---
title: "Summary Statistics - Wildfires and wells data"
output:
  html_notebook:
    theme: flatly
    highlight: haddock 
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: true
---

```{r setup, include=FALSE}
source("../../code/0-setup/01-setup.R")
library("lubridate")
```

```{r data_input, include=FALSE}
wells_all <- readRDS("../../data/interim/wells_all.rds") %>% 
  filter(year(date_latest) <= 2019 | is.na(date_latest))
wildfires_all <- readRDS("../../data/interim/wildfires_all.rds") %>% 
  filter(year %in% c(1984:2019))
```


# Wildfires data

### Summary stats

 - n = `r length(unique(wildfires_all$wildfire_id))` unique wildfires

**Mean wildfire area:**

```{r echo=FALSE}
####### pick up here; do we need this?
# wildfires_sf <- wildfires_all %>% st_as_sf(crs = crs_nad83) ##### change to WGS84
# wildfires_sf <- wildfires_sf %>% 
#   st_make_valid() %>% 
#   # group_by(wildfire_id) %>% 
#   # summarize(geometry = st_union(geometry)) %>% 
#   mutate(area = st_area(wildfires_sf)) ####
#   ##### adapted from redlining study:
#   #mutate(area_sqkm = (as.numeric(st_area(wildfires_all)) / 1000000))
```



### Plots

#### Plot. Wildfires by state

```{r echo=FALSE}
wildfires_all %>% 
  distinct(wildfire_id, .keep_all = TRUE) %>% 
  ggplot() +
  geom_bar(aes(state)) +
  theme_classic()
```

#### Plot. Wildfires by year

```{r echo=FALSE}
wildfires_all %>% 
  distinct(wildfire_id, .keep_all = TRUE) %>%
  ggplot() +
  geom_bar(aes(year)) +
  theme_classic()
```

#### Plot. Wildfires by year by state

```{r echo=FALSE}
wildfires_all %>% 
  distinct(wildfire_id, .keep_all = TRUE) %>%
  ggplot() +
  geom_bar(aes(year)) +
  facet_wrap(~ state) +
  theme_classic()
```


# Wells data

- There were `r wells_all %>% nrow()` wells in the analytic dataset
- Of these, `r wells_all %>% drop_na(date_earliest) %>% nrow()` had at least one operational date


### All wells

```{r echo=FALSE}
summary(wells_all)
```

### Wells with dates

```{r}
print("Average duration of operation (years):")
d <- wells_all %>% 
  drop_na(first_prod_date) %>% 
  drop_na(last_prod_date) %>% 
  mutate(prod_duration = last_prod_date - first_prod_date) %>% 
  select(prod_duration)
d <- d %>% mutate(prod_duration = as.numeric(prod_duration)) %>%  
  filter(prod_duration > 0 & prod_duration < 32689)  # known max; removes errors
summary(as.numeric(d$prod_duration) / 365.25)
sd(as.numeric(d$prod_duration) / 365.25)
```


