---
title: "Results - Wells by state and wildfire-year (primary analysis)"
output:
  html_notebook:
    theme: flatly
    highlight: haddock 
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
source("../../code/0-setup/01-setup.R")
```

```{r data_input, include=FALSE}
pop_exposed_state_year <- 
  read_csv("../../data/processed/pop_exposed_state_year.csv")
```

# Description

These analyses are for the assessments of the population exposed to wells in wildfire burn areas by state and year (1984-2029). For purposes of this analysis, we defined exposure as residing within 1 km of a well that was located in a wildfire burn area.

# Aggregated results for all states

## Population exposed to wells in wildfire burn areas

### Summary of population exposed by year

```{r include=TRUE}
d <- pop_exposed_state_year %>% 
  group_by(year) %>% 
  summarize(pop_exposed_n = sum(pop_exposed_n))
summary(d$pop_exposed_n)
print(paste0("SD = ", round(sd(d$pop_exposed_n), digits = 1)))
```

### Total population exposed by decade

-   1984-1989: **`r d <- pop_exposed_state_year %>% filter(year %in% c(1984:1989)) %>% select(pop_exposed_n) %>% sum(); round(d, digits = 0)`**
-   1990-1999: **`r d <- d <- pop_exposed_state_year %>% filter(year %in% c(1990:1999)) %>% select(pop_exposed_n) %>% sum(); round(d, digits = 0)`**
-   2000-2009: **`r pop_exposed_state_year %>% filter(year %in% c(2000:2009)) %>% select(pop_exposed_n) %>% sum(); round(d, digits = 0)`**
-   2010-2019: **`r d <- pop_exposed_state_year %>% filter(year %in% c(2010:2019)) %>% select(pop_exposed_n) %>% sum(); round(d, digits = 0)`**

### Average annual population exposed by decade

-   1984-1989: **`r d <- pop_exposed_state_year %>% filter(year %in% c(1984:1989)) %>% select(pop_exposed_n) %>% sum() / 6; round(d, digits = 1)` ± `r d <- pop_exposed_state_year %>% filter(year %in% c(1984:1989)) %>% select(pop_exposed_n); round(sd(d$pop_exposed_n), digits = 1)`**  
-   1984-1993: **`r d <- pop_exposed_state_year %>% filter(year %in% c(1984:1993)) %>% select(pop_exposed_n) %>% sum() / 6; round(d, digits = 1)` ± `r d <- pop_exposed_state_year %>% filter(year %in% c(1984:1993)) %>% select(pop_exposed_n); round(sd(d$pop_exposed_n), digits = 1)`**  
-   1990-1999: **`r d <- pop_exposed_state_year %>% filter(year %in% c(1990:1999)) %>% select(pop_exposed_n) %>% sum() / 10; round(d, digits = 1)` ± `r d <- pop_exposed_state_year %>% filter(year %in% c(1990:1999)) %>% select(pop_exposed_n); round(sd(d$pop_exposed_n), digits = 1)`**
-   2000-2009: **`r d <- pop_exposed_state_year %>% filter(year %in% c(2000:2009)) %>% select(pop_exposed_n) %>% sum() / 10; round(d, digits = 1)` ± `r d <- pop_exposed_state_year %>% filter(year %in% c(2000:2009)) %>% select(pop_exposed_n); round(sd(d$pop_exposed_n), digits = 1)`**
-   2010-2019: **`r d <- pop_exposed_state_year %>% filter(year %in% c(2010:2019)) %>% select(pop_exposed_n) %>% sum() / 10; round(d, digits = 1)` ± `r d <- pop_exposed_state_year %>% filter(year %in% c(2010:2019)) %>% select(pop_exposed_n); round(sd(d$pop_exposed_n), digits = 1)`**

### Plot. Population exposed by year

```{r echo=FALSE}
pop_exposed_state_year %>% 
  group_by(year) %>% 
  summarize(n = sum(pop_exposed_n)) %>% 
  ggplot() +
  geom_bar(aes(year, n), stat = "identity") +
  labs(x = "Year", y = "Population exposed (n)") +
  theme_classic()
```

### Table. Population exposed by year

```{r}
d <- pop_exposed_state_year %>% 
  group_by(year) %>% 
  summarize(n = round(sum(pop_exposed_n), digits = 0))
write_csv(d, "../../output/results/pop_exposed_state_year.csv")
d
```

### Plot. Population exposed by year with best fit line

Blue line is linear model; black line is cubic polynomial model

```{r echo=FALSE}
pop_exposed_state_year %>% 
  group_by(year) %>% 
  summarize(n = sum(pop_exposed_n)) %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  geom_smooth(method = "lm", color = "black", formula = y ~ poly(x, 3)) +
  labs(x = "Year", y = "Population exposed (n)") +
  theme_classic()
```

### Model. Population exposed in burn area by state-year

```{r}
model_fit <- glm(pop_exposed_n ~ year + factor(state),
                 data = pop_exposed_state_year)
summary(model_fit)
```

In a linear model adjusted for state, the exposed population increased significantly throughout the study period (p \< 0.005).

# Results stratified by state

## Population exposed by state

### Table. Population exposed by state

```{r echo=FALSE}
pop_exposed_state_year %>% 
  group_by(state) %>% 
  summarize(pop_exposed_n_by_state = round(sum(pop_exposed_n), digits = 0)) %>% 
  arrange(desc(pop_exposed_n_by_state))
```

### Plot. Population exposed by state

```{r echo=FALSE}
pop_exposed_state_year %>% 
  group_by(state) %>% 
  summarize(n = sum(pop_exposed_n)) %>% 
  ggplot() +
  geom_bar(aes(state, n), stat = "identity") +
  labs(x = "State", y = "Population exposed (n)") +
  theme_classic()
```

### Table. Population exposed by state and year

```{r}
pop_exposed_state_year %>% 
  group_by(state, year) %>% 
  summarize(n = round(sum(pop_exposed_n), digits = 0))
```

### Plot. Population exposed by state and year

```{r echo=FALSE}
pop_exposed_state_year %>% 
  group_by(state, year) %>% 
  summarize(n = sum(pop_exposed_n)) %>% 
  ggplot() +
  geom_bar(aes(year, n), stat = "identity") +
  labs(x = "Year", y = "Population exposed (n)") +
  facet_wrap(. ~ state) +
  theme_classic()
```

### Plot. Population exposed by year with best fit line

```{r echo=FALSE}
pop_exposed_state_year %>% 
  group_by(state, year) %>% 
  summarize(n = sum(pop_exposed_n)) %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(x = "Year", y = "Population exposed (n)") +
  facet_wrap(. ~ state) +
  theme_classic()
```

## Investigating time trends by state

Fit OLS linear regression models to look at the association between n wells in burn areas and time (in years). Some notable heterogeneity among the three states with the most wells in burn areas.

### Model. Population exposed by year and state (CA)

```{r}
model_fit <- glm(pop_exposed_n ~ year, family = "poisson",
                 data = subset(pop_exposed_state_year, state == "CA"))
summary(model_fit)
```

Weak association between advancing time and population exposed in California (p = 0.095)

### Model. Population exposed by year and state (OK)

```{r}
model_fit <- glm(pop_exposed_n ~ year,
                 data = subset(pop_exposed_state_year, state == "OK"))
summary(model_fit)
```

Increasing population exposure in Oklahoma (p \< 0.01).

### Model. Population exposed by year and state (TX)

```{r}
model_fit <- glm(pop_exposed_n ~ year,
                 data = subset(pop_exposed_state_year, state == "TX"))
summary(model_fit)
```

Weak association between advancing time and population exposed in Texas (p = 0.077)

### Model. Population exposed by year and state (KS)

```{r}
model_fit <- glm(pop_exposed_n ~ year,
                 data = subset(pop_exposed_state_year, state == "KS"))
summary(model_fit)
```

Positive time trend in KS (p = 0.001)

### Model. Population exposed by year and state (NM)

```{r}
model_fit <- glm(pop_exposed_n ~ year,
                 data = subset(pop_exposed_state_year, state == "NM"))
summary(model_fit)
```

No time trend in NM

### Model. Population exposed by year and state (AR)

```{r}
model_fit <- glm(pop_exposed_n ~ year,
                 data = subset(pop_exposed_state_year, state == "AR"))
summary(model_fit)
```

No time trend in AR
